---
title: "Exercise 2_2: Basic Raster Operations"
subtitle: 'Advanced Geospatial Data Processing for Social Scientists'
author: Dennis Abel & Stefan Jünger
format: 
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
editor: visual
callout-icon: false
callout-appearance: minimal
editor_options: 
  chunk_output_type: console
---

In the next three exercises, we’ll dive into the world of raster data.
You'll practice importing different raster formats, explore the properties of spatial layers, and even create a new raster layer by combining datasets.

Along the way, you might encounter a few common challenges, like dealing with different coordinate systems. These are great opportunities to sharpen your spatial data skills! 🚀🗺️

Let’s jump in!

### Exercises

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 1

In this exercise, you’ll import a raster `.tif` file from the `./data/z22` folder.
Once loaded, take a moment to explore the raster's dimensions, resolution, and other key properties.
Feel free to visualize it to get an initial sense of its content!

:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

You can use the `terra` package to load raster files. With the `terra::rast()` function, you’ll be able to load your `.tif` file and start working with it.

:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 2

In this exercise, you’ll import a NetCDF file (`.nc`) from the `./data/z22` folder.
Once you’ve successfully loaded it, check out the dimensions and variables within the file.
This is a great chance to get familiar with different raster data file formats!

:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

To work with `.nc` files, you can also use the `terra` package. The `terra::rast()` function works for NetCDF files, just like for raster TIF files.

:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 3

Now it’s time to get creative! 🎨 You’ve already imported both a TIF file and a NetCDF file. In this exercise, combine these two datasets to create a new raster layer. You can experiment by performing basic operations like adding, multiplying, or averaging the values from the two layers.

:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip
However, beware:
Since the CRS of the two datasets might not match, you could encounter a warning.
Take a moment to investigate and correct it!

- Check the CRS of both datasets with `terra::crs()`.
- If necessary, use `terra::project()` to align the CRS of both layers.
- Once aligned, you can perform mathematical operations like +, -, *, or / on the two layers.

:::

### Solutions

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 1

```{r}
#| eval: false
library(terra)

# Load the raster data
age_65_tif <- terra::rast("./data/z22/age_from_65.tif")

# Explore the raster
print(age_65_tif)                 # Check basic info
dim(age_65_tif)                   # Check dimensions
terra::res(age_65_tif)            # Resolution

# Optional: visualize it
plot(age_65_tif)
```

```{r}
#| echo: false
library(terra)

# Load the raster data
age_65_tif <- terra::rast("../../data/z22/age_from_65.tif")

# Explore the raster
print(age_65_tif)                 # Check basic info
dim(age_65_tif)                   # Check dimensions
terra::res(age_65_tif)            # Resolution

# Optional: visualize it
plot(age_65_tif)
```

:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 2

```{r}
#| eval: false
# Load the NetCDF file
population_nc <- terra::rast("./data/z22/population.nc")

# Explore the NetCDF file
print(population_nc)             # Check basic info
dim(population_nc)               # View dimensions
names(population_nc)             # Check variable names

# Optional: visualize one of the variables
plot(population_nc[[1]])         # Plot the first variable
```

```{r}
#| echo: false
# Load the NetCDF file
population_nc <- terra::rast("../../data/z22/population.nc")

# Explore the NetCDF file
print(population_nc)             # Check basic info
dim(population_nc)               # View dimensions
names(population_nc)             # Check variable names

# Optional: visualize one of the variables
plot(population_nc[[1]])         # Plot the first variable
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 3

```{r}
# Check the CRS of both datasets
crs_age_65_tif <- terra::crs(age_65_tif)
crs_population_nc <- terra::crs(population_nc)

# If CRS don't match, align them
if (crs_age_65_tif != crs_population_nc) {
  age_65_tif <- terra::project(age_65_tif, population_nc)
}

# Combine the layers (example: calculate number of people >= 65 years)
old_peeps <- population_nc * (age_65_tif / 100)

# Explore the new raster layer
plot(old_peeps)
```

:::