---
title: "Exercise 4_1: Simple Map"
subtitle: 'Advanced Geospatial Data Processing for Social Scientists'
author: Dennis Abel & Stefan Jünger
format: 
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
editor: visual
callout-icon: false
callout-appearance: minimal
---

### Exercises

We are now focussing on creating a first map based on raster data in `ggplot`. We will work with population data in the US.

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 1

Load the WorldPop file for the USA in 2020 (`./data/usa_ppp_2020_1km_Aggregated.tif`). Create a simple plot with `terra::plot()` to inspect the data.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 2

We want to focus on the 48 contiguous States and exclude Alaska, Hawaii as well as overseas territories. To do that, please:

-   load a US States shapefile with the [tigris](https://github.com/walkerke/tigris) package,
-   filter it to the 48 contiguous states,
-   and plot the shapefile to cross-check your progress.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

We haven't worked with `tigris` before. Below you find the code to load and select the relevant US states.

```{r}
#| eval: false
# install.packages("tigris")
library(tigris)

US_states <- states() |> 
  dplyr::filter(!STUSPS %in% c("AK", "AS", "GU", "HI", "MP", "PR", "VI"))
```
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 3

We now want to crop the raster-file on the basis of the US states bounding box: - Create a bounding box based on your shapefile, - Reproject this bbox to the CRS of the raster file, - Then crop the raster file on based on the bbox.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

You will need three `sf`-commands to prepare your bbox: `st_bbox()` to create the bbox, `st_as_sfc()` to transform it into an `sf`-object and `st_transform()` to reproject it to the CRS of your raster file. The code below shows the entire pipe for these steps.

```{r}
#| eval: false
bbox <- sf::st_bbox(US_states) |> 
  sf::st_as_sfc() |> 
  sf::st_transform(crs = terra::crs(USA_pop_2020))
```
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 4

Inspect the attributes of your raster file. The CRS is currently in WGS84. Is that suitable for a plot of the US territory? Alternatively, "NAD83 / Conus Albers" is often used for data analysis and presentation of the 48 contiguous US states. It is an equal-area projection and minimizes distortion across the US territory.

-   Search for the EPSG of "NAD83 / Conus Albers" on <https://spatialreference.org/ref/>,
-   Reproject the map and your shapefile accordingly.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 5

Now we can plot the population data together with the shapefile for the US states. Adjust color schemes accordingly and add labels.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

Unfortunately, our data is right-skewed and contains a few outliers with high population numbers per grid cell. This is a challenge for our color-coding. When working with `scale_fill_distiller` or `scale_fill_viridis_c` to adjust your color scheme, you could utilize the arguments `limits` and `oob = scales::squish` to account for that.

```{r}
#| eval: false
ggplot()+
  ... +
  scale_fill_distiller(...,
                       limits = c(0, 1000), 
                       oob = scales::squish
                       ) +
  ...
```
:::

### Solutions

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 1

```{r}
#| eval: false
# load libraries
library(dplyr)
library(sf)
library(ggplot2)
library(terra)
library(tidyterra)

# Load the data
USA_pop_2020 <- terra::rast("./data/usa_ppp_2020_1km_Aggregated.tif")

# Inspect it
terra::plot(USA_pop_2020)

```

```{r}
#| echo: false
# load libraries
library(dplyr)
library(sf)
library(ggplot2)
library(terra)
library(tidyterra)

# Load the data
USA_pop_2020 <- terra::rast("../../data/usa_ppp_2020_1km_Aggregated.tif")

# Inspect it
terra::plot(USA_pop_2020)

```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 2

```{r}
# install.packages("tigris")
library(tigris)

# Load the states and filter out non-contiguous territories
US_states <- states() |> 
  dplyr::filter(!STUSPS %in% c("AK", "AS", "GU", "HI", "MP", "PR", "VI"))

plot(sf::st_geometry(US_states))
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 3

```{r}
bbox <- sf::st_bbox(US_states) |> 
  sf::st_as_sfc() |> 
  sf::st_transform(crs = terra::crs(USA_pop_2020))

USA_pop_2020 <- terra::crop(
  USA_pop_2020,
  bbox
)
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 4

```{r}
USA_pop_2020 <- terra::project(USA_pop_2020, "EPSG:5070")
US_states <- sf::st_transform(US_states, crs = "EPSG:5070")
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 5

```{r}
ggplot()+
  tidyterra::geom_spatraster(data = USA_pop_2020)+
  geom_sf(data = US_states,
          fill = "transparent",
          color = "black",
          size = 5) +
  scale_fill_distiller(type = "seq", 
                       palette = "PuRd", 
                       direction = 1,
                       na.value = "transparent",
                       limits = c(0, 1000), 
                       oob = scales::squish
                       ) +
  labs(
    fill = "Population\ncount",
    title = "Estimated population of the USA in 2020",
    subtitle = "Approx. 1x1km grid"
  )
```
:::
