---
title: "Exercise 4_2: Fancy Map"
subtitle: 'Advanced Geospatial Data Processing for Social Scientists'
author: Dennis Abel & Stefan Jünger
format: 
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
editor: visual
callout-icon: false
callout-appearance: minimal
---

### Exercises

In our first mapping exercise, we have created a map of the population distribution in the US in 2020. We now want to create a second graph which zooms into Texas. Instead of visualizing the population at one point in time, we are aiming at describing changes in population between 2000 and 2020.

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 1

What data we require: We need the two population raster layers for Texas in 2000 and 2020 (`./data/US-TX_ppp_2000_1km.tif` and `./data/US-TX_ppp_2020_1km.tif`). We also need county shapefiles for Texas, which can be downloaded with the `tigris` package (`tigris::counties("TX")`). 

After loading the three objects, create a simple plot of the counties shapefile.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 2

In order to study population dynamics in Texas, let's calculate simple differences between 2000 and 2020 for each grid cell based on our two raster layers for Texas. Store the result in a new raster file and visualize a simple plot of it. 

Texas is one of the US states with the highest population increase in the last decades - this should be reflected in the map.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 3

Before we combine our new raster file with our counties shapefile, we want our raster and vector data to have matching CRS. "NAD83 / Texas Centric Mapping System Albers Equal Area" is specifically projected for Texas.

-   Search for the EPSG of "NAD83 / Texas Centric Mapping System Albers Equal Area" on <https://spatialreference.org/ref/>,
-   Reproject the raster file which stores the population changes and your counties shapefile accordingly.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 4

Let's use `ggplot` to create a plot which contains the shapefile and raster file. Adjust the color scheme by utilizing `scale_fill_viridis_c`, `scale_fill_distiller`, or `scale_fill_gradient2`.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 5

We want to point out counties which experience high population growth. Please attach labels to the top three counties by population growth. You can utilize `slice_head(n = 3)` after ordering the data in descending order to select only those three observations. Add this as an additional layer to the plot as text or text labels.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 6

What else could enhance your map? Polish it for publication, for example, by:

- Turn NAs transparent, 
- Add titles, 
- Adjust the overall theme of the plot, 
- Add a North arrow and scale bar.
:::

### Solutions

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 1

```{r}
#| eval: false
library(dplyr)
library(sf)
library(ggplot2)
library(terra)
library(tidyterra)
library(tigris)

# Load the raster data
TX_pop_2020 <- terra::rast("./data/US-TX_ppp_2020_1km.tif")
TX_pop_2000 <- terra::rast("./data/US-TX_ppp_2000_1km.tif")

# Load Texas county data
TX_counties <- tigris::counties("TX", progress_bar = FALSE)

plot(st_geometry(TX_counties))

```

```{r}
#| echo: false
library(dplyr)
library(sf)
library(ggplot2)
library(terra)
library(tidyterra)
library(tigris)

# Load the raster data
TX_pop_2020 <- terra::rast("../../data/US-TX_ppp_2020_1km.tif")
TX_pop_2000 <- terra::rast("../../data/US-TX_ppp_2000_1km.tif")

# Load Texas county data
TX_counties <- tigris::counties("TX", progress_bar = FALSE)

plot(st_geometry(TX_counties))

```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 2

```{r}
# Let's calculate the deviation
TX_pop_diff <- TX_pop_2020 - TX_pop_2000

TX_pop_diff

terra::plot(TX_pop_diff)
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 3

```{r}
TX_pop_diff <- terra::project(TX_pop_diff, "EPSG:3081")
TX_counties <- sf::st_transform(TX_counties, crs = "EPSG:3081")
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 4

```{r}
# Let's adjust color palette to account for diverging values
ggplot()+
  geom_spatraster(data = TX_pop_diff)+
  geom_sf(data = TX_counties,
          fill = "transparent",
          color = "white",
          size = 3)+
  scale_fill_viridis_c(option = "magma")

# Identify max and min to adjust for skewness of positive and negative values
max_val <- max(abs(minmax(TX_pop_diff)))

ggplot()+
  geom_spatraster(data = TX_pop_diff)+
  geom_sf(data = TX_counties,
          fill = "transparent",
          color = "black",
          size = 3)+
  scale_fill_distiller(type = "div", palette = "PuOr", limits = c(-max_val, max_val))

# Or define manually for even greater control
ggplot()+
  geom_spatraster(data = TX_pop_diff)+
  geom_sf(data = TX_counties,
          fill = "transparent",
          color = "black",
          size = 3)+
  scale_fill_gradient2(
    low = "#7f3b08",
    mid = "white",
    high = "#40004b",
    midpoint = 0
  )
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 5

```{r}
# Identify growing and shrinking counties
TX_counties <- TX_pop_diff |> 
  extract(TX_counties, fun = mean, na.rm =TRUE) |> 
  as_tibble() %>% 
  bind_cols(TX_counties, .)

# Select top three counties by population growth
extremes <- TX_counties |> 
  arrange(desc(`US-TX_ppp_2020_1km`)) |> 
  slice_head(n=3)

# Add geom_sf_text layer to the plot 
ggplot()+
  geom_spatraster(data = TX_pop_diff)+
  geom_sf(data = TX_counties,
          fill = "transparent",
          color = "black",
          size = 3)+
  geom_sf_text(data = extremes, aes(label = NAMELSAD), size = 4, color = "black")+ 
  scale_fill_gradient2(
    low = "#7f3b08",
    mid = "white",
    high = "#40004b",
    midpoint = 0
  )
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 6

```{r}
# Final polishing
library(ggspatial)
library(ggrepel)

ggplot()+
  geom_spatraster(data = TX_pop_diff)+
  geom_sf(data = TX_counties,
          fill = "transparent",
          color = "black",
          size = 3)+
  geom_label_repel(
    data = extremes, 
    aes(geometry = geometry, label = NAMELSAD), 
    stat = "sf_coordinates",
    min.segment.length = 0, 
    segment.color = "black",   
    segment.size = 0.5,        
    size = 4,                  
    fill = "white",            
    color = "black",           
    label.size = 0.2           
  )+ 
  scale_fill_gradient2(
    low = "#7f3b08",
    mid = "white",
    high = "#40004b",
    midpoint = 0,
    na.value = "transparent",
    name = "Population\nChange"
  )+
  theme_grey()+
  theme(
    axis.title = element_blank()
  ) +
  labs(
    title = "Population change in Texas counties (2000–2020)",
    subtitle = "In absolute numbers on 1x1km grid\nText labels identify top three districts",
    caption = "Source: WorldPop (2018)"
  ) +
  annotation_scale(
    location = "br",
    width_hint = 0.3
  ) +
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    style = north_arrow_fancy_orienteering
  )
```
:::
