---
title: "Exercise 5_2: Raster cube with stars"
subtitle: 'Advanced Geospatial Data Processing for Social Scientists'
author: Dennis Abel & Stefan Jünger
format: 
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
editor: visual
callout-icon: false
callout-appearance: minimal
---

### Exercises

We are deliberately boring in this exercise - we want to do exactly the same as before - this time with `stars`. Our aim is to create a raster cube (`stars`) for Texas population data between 2017 and 2020.

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 1

Load the four population raster layers for Texas 2017-2020 (`./data/US-TX_ppp_2017_1km.tif`-`./data/US-TX_ppp_2020_1km.tif`) with `stars::read_stars()` and integrate these into one raster cube. First, create a file list for all four raster layers. Second, prepare a vector which stores the four years. Finally, you can read in the file list along the list of years.

:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 2

Adjust the attribute name (`names()`) to a meaningful string for your variable.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 3

Again, let's reproject our Texas data to EPSG:3081. Inspect the object afterwards (`print()`).
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 4

Visualize all four years with a spatial window of `x = 0:1000` and `y = 0:1000`. You can input your `stars` object into base `R` `plot()` command. The default color scale is often insufficient. Specify the argument `breaks = "equal"` to improve readability.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 5

Store your `stars` object as `.rds` file on disk (`saveRDS`).
:::

### Solutions

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 1

```{r}
#| eval: false
library(dplyr)
library(stars)

# Load the raster data
files <- list.files("./data", pattern = "US-TX_ppp_20(17|18|19|20)_1km\\.tif$", full.names = TRUE)

# Vector of years
dates <- 2017:2020

# Read layers along time dimension
TX_pop_cube <- read_stars(files, along = list(time = dates))

print(TX_pop_cube)
```

```{r}
#| echo: false
library(dplyr)
library(stars)

# Load the raster data
files <- list.files("../../data", pattern = "US-TX_ppp_20(17|18|19|20)_1km\\.tif$", full.names = TRUE)

# Vector of years
dates <- 2017:2020

# Read layers along time dimension
TX_pop_cube <- read_stars(files, along = list(time = dates))

print(TX_pop_cube)
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 2

```{r}
names(TX_pop_cube) <- "population"

st_dimensions(TX_pop_cube)

print(TX_pop_cube)
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 3

```{r}
grid <- st_bbox(TX_pop_cube) |>
  st_as_sfc() |>
  st_transform(crs = 3081) |>
  st_bbox() |>
  st_as_stars(nx = dim(TX_pop_cube)["x"], 
              ny = dim(TX_pop_cube)["y"])

TX_pop_cube <- st_warp(TX_pop_cube, grid)

print(TX_pop_cube)
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 4

```{r}
# Plot a single layer
plot(TX_pop_cube[1,
                 0:1000,
                 0:1000, 
                 1:4],
     breaks = "equal")
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 5

```{r}
#| eval: false
# Export
saveRDS(TX_pop_cube, "./data/CA_pop_cube.rds")
```

```{r}
#| echo: false
# Export
saveRDS(TX_pop_cube, "../../data/CA_pop_cube.rds")
```
:::
