---
title: "Exercise 6_1: A Map with many facets"
subtitle: 'Advanced Geospatial Data Processing for Social Scientists'
author: Dennis Abel & Stefan Jünger
format: 
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
editor: visual
callout-icon: false
callout-appearance: minimal
---

### Exercises

Now that we know how to map several layers simultaneously with `ggplot()`, we want to explore the population in Texas in the four years 2017-2020. We will plot both `terra` and `stars` objects.

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 1

In the previous session, we have created a raster stack and a raster cube for the four years in Texas. If these are not in your environment anymore, please load both (`./data/TX_pop_stack.tif` and `./data/TX_pop_cube.rds`). 

In addition, load the county shapefiles for Texas with the `tigris` package. Reproject it to the CRS of your raster files.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 2

You are already proficient in creating publication-ready maps. Please create two plots - one for the `terra` object and another one for the `stars` object. Utilize the `facet_wrap()` command to separate the plot along the four years of the data.

Polish it as much as you like (NAs transparent, theme, titles, and North arrow and scale bar,...).
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 3

Add your Texas counties shapefile to the plots with `geom_sf()`. 
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 4

Let's use `ggplot` to create a plot which contains the shapefile and raster file. Adjust the color scheme by utilizing `scale_fill_viridis_c`, `scale_fill_distiller`, or `scale_fill_gradient2`.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 5

We want to point out counties which experience high population growth. Please attach labels to the top three counties by population growth. You can utilize `slice_head(n = 3)` after ordering the data in descending order to select only those three observations. Add this as an additional layer to the plot as text or text labels.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 6

What else could enhance your map? Polish it for publication, for example, by:

-   Turn NAs transparent,
-   Add titles,
-   Adjust the overall theme of the plot,
-   Add a North arrow and scale bar.
:::

### Solutions

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 1

```{r}
#| eval: false
library(dplyr)
library(sf)
library(ggplot2)
library(terra)
library(tidyterra)
library(tigris)
library(stars)

# Load the raster data
TX_pop_stack <- terra::rast("./data/TX_pop_stack.tif")

# Load the raster cube
TX_pop_cube <- readRDS("./data/TX_pop_cube.rds")

# Load Texas county data
TX_counties <- tigris::counties("TX", progress_bar = FALSE)

crs_raster <- terra::crs(TX_pop_stack, describe = TRUE)$code

TX_counties <- TX_counties |> 
  sf::st_transform(crs = as.numeric(crs_raster))
```

```{r}
#| echo: false
library(dplyr)
library(sf)
library(ggplot2)
library(terra)
library(tidyterra)
library(tigris)
library(stars)

# Load the raster data
TX_pop_stack <- terra::rast("../../data/TX_pop_stack.tif")

# Load the raster cube
TX_pop_cube <- readRDS("../../data/TX_pop_cube.rds")

# Load Texas county data
TX_counties <- tigris::counties("TX", progress_bar = FALSE)

crs_raster <- terra::crs(TX_pop_stack, describe = TRUE)$code

TX_counties <- TX_counties |> 
  sf::st_transform(crs = as.numeric(crs_raster))
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 2

```{r}
library(ggspatial)

ggplot()+
  geom_spatraster(data = TX_pop_stack)+
  scale_fill_viridis_c(na.value = "transparent",
                       name = "Pop count",
                       limits = c(0, 2000),
                       oob = scales::squish)+
  facet_wrap(~ lyr)+
  theme_minimal()+
  labs(
    title = "Population in Texas",
    subtitle = "In absolute numbers on 1x1km grid",
    caption = "Source: WorldPop (2018)"
  ) +
  annotation_scale(
    location = "bl",
    width_hint = 0.3
  ) +
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    style = north_arrow_fancy_orienteering
  )
```

```{r}
library(ggspatial)

ggplot() +
  geom_stars(data = TX_pop_cube, 
             aes(fill = population)
             ) +
  scale_fill_viridis_c(na.value = "transparent",
                       name = "Pop count",
                       limits = c(0, 2000),
                       oob = scales::squish)+
  coord_equal() +
  facet_wrap(~ time)+
  theme_minimal()+
  labs(
    title = "Population in Texas",
    subtitle = "In absolute numbers on 1x1km grid",
    caption = "Source: WorldPop (2018)"
  ) +
  annotation_scale(
    location = "bl",
    width_hint = 0.3
  ) +
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    style = north_arrow_fancy_orienteering
  )
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 3

```{r}
library(ggspatial)

ggplot()+
  geom_spatraster(data = TX_pop_stack)+
  geom_sf(data = TX_counties,
          fill = "transparent",
          color = "white",
          size = 3)+
  scale_fill_viridis_c(na.value = "transparent",
                       name = "Pop count",
                       limits = c(0, 2000),
                       oob = scales::squish)+
  facet_wrap(~ lyr)+
  theme_minimal()+
  labs(
    title = "Population in Texas",
    subtitle = "In absolute numbers on 1x1km grid",
    caption = "Source: WorldPop (2018)"
  ) +
  annotation_scale(
    location = "bl",
    width_hint = 0.3
  ) +
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    style = north_arrow_fancy_orienteering
  )
```

```{r}
library(ggspatial)

ggplot() +
  geom_stars(data = TX_pop_cube, 
             aes(fill = population)
             ) +
  geom_sf(data = TX_counties,
          fill = "transparent",
          color = "white",
          size = 3)+
  scale_fill_viridis_c(na.value = "transparent",
                       name = "Pop count",
                       limits = c(0, 2000),
                       oob = scales::squish)+
  coord_sf() +
  facet_wrap(~ time)+
  theme_minimal()+
  labs(
    title = "Population in Texas",
    subtitle = "In absolute numbers on 1x1km grid",
    caption = "Source: WorldPop (2018)"
  ) +
  annotation_scale(
    location = "bl",
    width_hint = 0.3
  ) +
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    style = north_arrow_fancy_orienteering
  )
```
:::


